A Technique for Inter-Catalogue Recommendations

### 2014-06-11 paper2/paper.tex
Finish reworking the LatentFeatureRecommender subsection.

Toward this end, we borrow the concept of {\em Term Frequency--Inverse Document
Frequency} (tfidf) from the information retrieval community. In using tfidf, we
are considering each item description to be a ``document''.  Thus we define the
tfidf of a term $t$ within an item description $d$ that's part of a combined
category catalogue $C$ (i.e. the catalogue consisting of all items in both
retailers' catalogues for a particular common category) as
\begin{align}
\mathrm{tfidf}(t,d,C) = f(t,d) \times
\mathrm{log}\frac{\left|C\right|}{\left|\left\{d \in C : t \in
d\right\}\right|},
\end{align}
where $f(t,d)$ is the number of occurances of term $t$ in item description $d$,
and $\left|\left\{d \in C : t \in d\right\}\right|$ is the number of items in
$C$ with descriptions containing at least one instance of term $t$.  We can now
represent each item as a sparse vector of tfidf values, which has a nonzero
element for each term in its description.

A key feature of tfidf is that terms appearing in many item descriptions are
discounted via the idf factor. Conversely, rare terms will have larger tfidf
values. These features help capture the most relevant words in each description.
We now define the $ContentSimilarity$ between items $A$ and $B$ as
\begin{align}
ContentSimilarity(A,B) = v_A \cdot v_B
\end{align}
where $v_A$ and $v_B$ are the tfidf vector representations of items $A$ and $B$,
respectively.

@@

The $ContentBasedRecommender$ deserves a few more comments. For starters, the
use of an inner product, which is bounded only by the length of the tfidf
vectors, is not fundamental to this algorithm. A bounded similarity function
such as cosine similarity could have been used instead, without qualitatively
changing the results discussed later in this report. This is because the lengths
of the product descriptions have little variation, and therefore, the vector
norms of the tfidf vectors have little variation. Finally, we mention that while
this algorithm is quadratic in the number of items, it can be executed relatively
efficiently using matrix multiplication, which trades time for space (i.e.
memory). Furthermore, confining our experiments to individual categories reduces
the impact of the quadratic time complexity. In our case, Macy's entire
catalogue consists of over $60,000$ products, making a quadratic time algorithm
challenging. However, the average category has about two thousand products,
which is significantly more manageable.

@@

We can begin to address these issues by leveraging the recommendations we
already have between products within a single partition. One approach is to
consider the {\em neighborhood} of each product, that is a product along with
each of its immediate neighbors in the recommendation graph. With this, we
create a new similarity function called $NeighborhoodSimilarity$. This function
uses the $ContentSimilarity$ function from before to compare every pair of
products between the two neighborhoods of a pair of products $A$ and $B$:
\begin{align}
NeighborhoodSimilarity(A, B) = \sum\limits_{S\in
Neigh(A)}\sum\limits_{T\in Neigh(B)}
ContentSimilarity(S, T) 
\end{align}
where $Neigh(A)$ is the set of items that share edges with item $A$. We use this
similarity function to construct the $NeighborhoodRecommender$, which is
an extension of the $ContentBasedRecommender$. The algorithm is the same as
before except now we choose a recommendation associated with items $A$ and $B$
by maximnizing
\begin{align}
ContentSimilarity(A,B) + \eta \cdot NeighborhoodSimilarity(A, B),
\end{align}
where $\eta$ is a parameter. We find that $\eta = 0.1$ yields good results for
this algorithm.

@@

\subsection*{\em LatentConceptRecommender}
Although the $NeighborhoodRecommender$ leverages the information contained in
the recommendation graphs to some extent, it is still primarily based on product
descriptions. Presumably, the graphs are derivative of a collaborative filtering
scheme, and hence, reveal associations between products that are not evident
from the language of their descriptions. For example, a dress's description
might not capture if it is more conservative or more risque. Yet, this could be
the salient feature it has in common with an associated recommendation.  These
sort of abstract, perhaps ineffable, features can be essential to item-item
relations. We must therefore ask whether we can more fully utilize the given
graphical information to discover these latent features in order to choose
better inter-catalogue associations.

@@

We may then construct a synthetic $N \times N$ user-item matrix $M$ with rows
representing items and columns representing users, which are in fact the
stationary distributions discovered from our random walks. This matrix has at
least one notable advantage over most real user-item matrices---it is not
sparse.  Rather, each synthetic user has a finite value associated with every
item that is reachable from its origin item in the graph. Unlike the typical CF
scenario, we may therefore employ SVD as our matrix factorization technique
since our user-item matrix is virtually complete (i.e. nearly void of empty
entries).  Using SVD, we factor our user-item matrix as
\begin{align}
M = U \Sigma V^T,
\end{align}
where the rows of the $N \times N$ matrix $U$ and the $N \times N$ matrix $V$
are the items and users represented in an $N$-dimensional concept-space,
respectively, and the $N$ diagonal entries of $\Sigma$ are the singular values
associated with each concept.

Having transformed our items into concept-space, we now discard all but the $k$
leftmost columns of the matrix $U$. The idea is that these leftmost columns are
the most significant eigenvectors of concept-space and capture most, if not all,
of the relevant latent features. Meanwhile, the discarded columns are presumed
to contain mostly noise. This rank-reduction process yields matrices that are
typically written as
\begin{align}
M \approx U_k \Sigma_k V^T_k,
\end{align}
where the rows of $U_k$ and $V_k$ are the items and users represented in a
$k$-dimensional concept-space, respectively.  Rank-reducing following the
application of SVD in this way is a common technique used in information
retrieval and is often referred to as {\em Latent Semantec Indexing} (LSI)
\cite{Deerwester1990}.

@@

If a mapping between the latent spaces associated with two catalogues were
known, we could translate all items into a common space and then choose
inter-catalogue edges between item pairs that are nearest to one another in
concept-space. To learn these mapping, we must revisit the notion of content
similarity. Once again, we utilize tf-idf vectors. This time, however, we weight
an item's tf-idf vector by the elements in the item's latent feature vector. In
this way, we construct a dense tf-idf vector for each concept for each graph
partition. Formally, we say that each concept $j$ is represented by 
\begin{align}
EXPRESSION FOR A CONCEPT'S TF-IDF VECTOR
\end{align}

In this case, we represent each concept (rather than each product) as a vector
of words. To represent a concept in terms of the vocabulary, we calculate the
value of each word by summing the tf-idf of that word for each product $P$,
weighted by the projection of $P$ on our concept in concept-space. That is:
\begin{align}
C_w = \sum\limits_{P} \text{tfidf}(w)\times P_C
\end{align}
where $C_w$ is the value of concept $C$'s representation in dimension $w$, the
sum is over all products $P$ in the catalgoue, tfidf($w$) is the tfidf score
for word $w$, and $P_C$ is the projection of $P$ on concpet $C$'s axis in
concept-space.

Now for a concept such as {\em formal}, the words ``sophisticated'' or
``official'' may have high value since products with those descriptive words
would also have a large amount of the {\em formal} concept in their
concept-representations. Using these vector representation, we calculate the
inner product between two concepts to find how similar they are to eachother.
This can be used to create a mapping from one set of concepts to another as it
captures the relative breakdown of a concept from one space to the concepts in
the other.

In all, the latent-concept approach consists of four phases. First, we simulate
user sessions by randomly walking the recommendation graphs. Second, we learn
concept spaces for products in each catalogue via SVD. Third, we find a mapping
between concept-spaces using the aggregated descriptions of all products. And
lastly, we find the nearest neighbors of each product in the concept-space of
the other catalogue concept-space. These nearest neighbors become the predicted
recommendations between products across the two catalogues. We call this method
the $LatentConceptRecommender$

@@

\subsection*{Concept Mapping}
A tricky operation of this procedure is the concept mapping. We attempt to find
the best linear transformation from one concept-space to the other. We do this
by representing each concept as a vector of words in the vocabulary, each
weighted with how well it represents the concept. We construct this by weighing
the tfidf of each word in a product's description by how well that word
represents the concept. Note that products can have a positive or negative value
for each concept, thus the word representation of concpets are also positive and
negative.

@@

One way to normalize is to normalize each column of the inner-product matrix by
the L1 norm. This preserves the L1 norm of products after the transformation,
but does not preserve the L2 norm. Similarly, we can normalize the columns by
the L2 norm. Neither of these methods preserve the L2 norm, which would be ideal
as we find the nearest neighbors in the other concept-space by their L2
distance.

@@

Proceeding as before, we find the nearest products in concept-space. However,
before deeming the nearest neighbors to be the best recommendations, we divide
the distance to each product by its popularity. Since we model distance in 
concept-space to be inversely proportional to similarity, this reincorporation
of popularity is analagous to multiplying similarity by popularity, as done in
the original construction of the recommendation graph.

@@

It is also informative to test the efficacy of the concept mapping, since this
such a hard task. As in product-product content-similarity, it is difficult to
capture similarity using descriptions alone. It is not even clear that a mapping 
necessarily exists between two seperately learned concept-spaces. The concepts 
may represent fundamentally different things, which cannot be well mapped to 
each other. We can measure the efficacy of the mapping by comparing it to a 
couple of baselines.

The first baseline is in fact an upper bound of sorts on the mapping. This
baseline recommender is called the One Model Recommender. Unlike the other
baselines, this recommender is not a valid one, as it cheats by using
information from the unpartitioned graph. The One Model Recommender learns a
latent concept-space over the entire recommendation graph by allowing the random
walkers to traverse it before it is partitioned. In every other respect, it is
the same as the LSI Recommender. However, since products from both catalogues
are represented in the same concept-space, there is no need for a mapping
between concept-spaces.

@@

\subsection*{Results}
[Number of products mapped to in different methods]
\begin{center}
\begin{tabular}{ | l | c | c | c |}
\hline
Normalization & 2-Precision & 2-Recall & 2-F1 Score \\ \hline\hline
L1 normalize each column of matrix &&&\\ \hline
L2 normalize each column of matrix &&&\\ \hline
Orthogonalize matrix &&&\\ \hline
Normalize products in concept-space &&&\\ \hline
\end{tabular}
\end{center}

>>>>

Toward this end, we borrow the concept of {\em Term Frequency--Inverse Document
Frequency} (tf-idf) from the information retrieval community. In using tf-idf, we
are considering each item description to be a ``document''.  Thus we define the
tf-idf of a term $t$ within an item description $d$ that's part of a combined
category catalogue $C$ (i.e. the catalogue consisting of all items in both
retailers' catalogues for a particular common category) as
\begin{align}
\textrm{tf-idf}(t,d,C) = f(t,d) \times
\mathrm{log}\frac{\left|C\right|}{\left|\left\{d \in C : t \in
d\right\}\right|},
\end{align}
where $f(t,d)$ is the number of occurances of term $t$ in item description $d$,
and $\left|\left\{d \in C : t \in d\right\}\right|$ is the number of items in
$C$ with descriptions containing at least one instance of term $t$.  We can now
represent each item as a sparse vector of tf-idf values, which has a nonzero
element for each term in its description.

A key feature of tf-idf is that terms appearing in many item descriptions are
discounted via the idf factor. Conversely, rare terms will have larger tf-idf
values. These features help capture the most relevant words in each description.
We now define the $ContentSimilarity$ between items $A$ and $B$ as
\begin{align}
ContentSimilarity(A,B) = \vec{\tau}(A) \cdot \vec{\tau}(B)
\end{align}
where $\vec{\tau}(A)$ is the tf-idf vector representation of item $A$.

@@

The $ContentBasedRecommender$ deserves a few more comments. For starters, the
use of an inner product, which is bounded only by the length of the tf-idf
vectors, is not fundamental to this algorithm. A bounded similarity function
such as cosine similarity could have been used instead, without qualitatively
changing the results discussed later in this report. This is because the lengths
of the product descriptions have little variation, and therefore, the vector
norms of the tf-idf vectors have little variation. Finally, we mention that
while this algorithm is quadratic in the number of items, it can be executed
relatively efficiently using matrix multiplication, which trades time for space
(i.e.  memory). Furthermore, confining our experiments to individual categories
reduces the impact of the quadratic time complexity. In our case, Macy's entire
catalogue consists of over $60,000$ products, making a quadratic time algorithm
challenging. However, the average category has about two thousand products,
which is significantly more manageable.

@@

We can begin to address these issues by leveraging the recommendations we
already have between products within a single partition. One approach is to
consider the {\em neighborhood} of each product, that is a product along with
each of its immediate neighbors in the recommendation graph. With this, we
create a new similarity function called $NeighborhoodSimilarity$. This function
uses the $ContentSimilarity$ function from before to compare every pair of
products between the two neighborhoods of a pair of products $A$ and $B$:
\begin{align}
NeighborhoodSimilarity(A, B) = \sum\limits_{i\in
Neigh(A)}\sum\limits_{j\in Neigh(B)}
ContentSimilarity(i, j) 
\end{align}
where $Neigh(A)$ is the set of items that share edges with item $A$. We use this
similarity function to construct the $NeighborhoodRecommender$, which is
an extension of the $ContentBasedRecommender$. The algorithm is the same as
before except now we choose a recommendation associated with items $A$ and $B$
by maximnizing
\begin{align}
ContentSimilarity(A,B) + \eta \cdot NeighborhoodSimilarity(A, B),
\end{align}
where $\eta$ is a parameter. We find that $\eta = 0.1$ yields good results for
this algorithm.

@@

\subsection*{\em LatentFeatureRecommender}
Although the $NeighborhoodRecommender$ leverages the information contained in
the recommendation graphs to some extent, it is still primarily based on product
descriptions. Presumably, the graphs are derivative of a collaborative filtering
scheme, and hence, reveal associations between products that are not evident
from the language of their descriptions. For example, a dress's description
might not capture if it is more conservative or more risque. Yet, this could be
the salient feature it has in common with an associated recommendation.  These
sort of abstract, perhaps ineffable, features can be essential to item-item
relations. We must therefore ask whether we can more fully utilize the given
graphical information to discover these latent features in order to choose
better inter-catalogue associations.

@@

We may then construct a synthetic $N \times N$ user-item matrix $M$ with rows
representing items and columns representing users, which are in fact the
stationary distributions discovered from our random walks. This matrix has at
least one notable advantage over most real user-item matrices---it is not
sparse.  Rather, each synthetic user has a finite value associated with every
item that is reachable from its origin item in the graph. Unlike the typical CF
scenario, we may therefore employ SVD as our matrix factorization technique
since our user-item matrix is virtually complete (i.e. nearly void of empty
entries).  Using SVD, we factor our user-item matrix as
\begin{align}
M = U \Sigma V^T,
\end{align}
where the rows of the $N \times N$ matrix $U$ and the $N \times N$ matrix $V$
are the items and users represented in an $N$-dimensional concept space,
respectively, and the $N$ diagonal entries of $\Sigma$ are the singular values
associated with each concept.

Having transformed our items into concept space, we now discard all but the $k$
leftmost columns of the matrix $U$. The idea is that these leftmost columns are
the most significant eigenvectors of concept space and capture most, if not all,
of the relevant latent features. Meanwhile, the discarded columns are presumed
to contain mostly noise. This rank-reduction process yields matrices that are
typically written as
\begin{align}
M \approx U_k \Sigma_k V^T_k,
\end{align}
where the rows of $U_k$ and $V_k$ are the items and users represented in a
$k$-dimensional concept space, respectively.  Rank-reducing following the
application of SVD in this way is a common technique used in information
retrieval and is often referred to as {\em Latent Semantec Indexing} (LSI)
\cite{Deerwester1990}.

@@

If a mapping between the latent spaces associated with two catalogues were
known, we could translate all items into a common space and then choose
inter-catalogue edges between item pairs that are nearest to one another in
concept space. To learn these mapping, we must revisit the notion of content
similarity. Once again, we utilize tf-idf vectors. This time, however, we weight
an item $i$'s tf-idf vector $\vec{\tau}(i)$ by the elements in its latent
feature vector $\vec{v}(i)$. In this way, we construct a dense {\em
super-tf-idf} vector for each concept for each catalogue. Formally, we say that
a concept $j$ of a catalogue $C$ is represented by the super-tf-idf vector
\begin{align}
\vec{s}(j,C) = \sum_{i \in C}{v(i)_j \cdot \vec{\tau}(i)},
\end{align}
where $v(i)_j$ is the $j$\textsuperscript{th} component of item $i$'s latent
feature vector. These super-tf-idf vectors act as a proxy for the concept. We
approximate the mapping between two concept spaces by taking inner products of
super-tf-idf vectors. The details of this mapping process are provided in a
subsequent section.

Together, these steps comprise the $LatentFeatureRecommender$. First, we
simulate user sessions by randomly walking the recommendation graphs. Second, we
learn concept spaces for products in each catalogue via LSI. Third, we find a
mapping between concept spaces using weighted tf-idf vector representations of
the products. And lastly, we find the nearest neighbors of each product from one
catalogue within the concept space of another catalogue. These nearest neighbors
are then the predicted recommendations between products across the two
catalogues.

%Now for a concept such as {\em formalness}, the words ``sophisticated'' or
%``classy'' may have high value since products with those descriptive words would
%also have a large amount of the {\em formalness} concept in their latent feature
%vectors.

@@

\subsection*{Concept Mapping}
A tricky operation of this procedure is the concept mapping. We attempt to find
the best linear transformation from one concept space to the other. We do this
by representing each concept as a vector of words in the vocabulary, each
weighted with how well it represents the concept. We construct this by weighing
the tf-idf of each word in a product's description by how well that word
represents the concept. Note that products can have a positive or negative value
for each concept, thus the word representation of concpets are also positive and
negative.

@@

One way to normalize is to normalize each column of the inner-product matrix by
the L1 norm. This preserves the L1 norm of products after the transformation,
but does not preserve the L2 norm. Similarly, we can normalize the columns by
the L2 norm. Neither of these methods preserve the L2 norm, which would be ideal
as we find the nearest neighbors in the other concept space by their L2
distance.

@@

Proceeding as before, we find the nearest products in concept space. However,
before deeming the nearest neighbors to be the best recommendations, we divide
the distance to each product by its popularity. Since we model distance in 
concept space to be inversely proportional to similarity, this reincorporation
of popularity is analagous to multiplying similarity by popularity, as done in
the original construction of the recommendation graph.

@@

It is also informative to test the efficacy of the concept mapping, since this
such a hard task. As in product-product content-similarity, it is difficult to
capture similarity using descriptions alone. It is not even clear that a mapping 
necessarily exists between two seperately learned concept spaces. The concepts 
may represent fundamentally different things, which cannot be well mapped to 
each other. We can measure the efficacy of the mapping by comparing it to a 
couple of baselines.

The first baseline is in fact an upper bound of sorts on the mapping. This
baseline recommender is called the One Model Recommender. Unlike the other
baselines, this recommender is not a valid one, as it cheats by using
information from the unpartitioned graph. The One Model Recommender learns a
latent concept space over the entire recommendation graph by allowing the random
walkers to traverse it before it is partitioned. In every other respect, it is
the same as the LSI Recommender. However, since products from both catalogues
are represented in the same concept space, there is no need for a mapping
between concept spaces.

@@

\subsection*{Results}
[Number of products mapped to in different methods]
\begin{center}
\begin{tabular}{ | l | c | c | c |}
\hline
Normalization & 2-Precision & 2-Recall & 2-F1 Score \\ \hline\hline
L1 normalize each column of matrix &&&\\ \hline
L2 normalize each column of matrix &&&\\ \hline
Orthogonalize matrix &&&\\ \hline
Normalize products in concept space &&&\\ \hline
\end{tabular}
\end{center}
